Powering Hidden Markov Model by Neural Network based Generative Models

### 2019-11-18 manuscript/main.tex
update content

Given the success of GMM-HMM, it is not efficient for modeling data in nonlinear manifold. Research attempts at training HMM with neural networks have been made to boost the modeling capacity of HMM. A successful work of this track has brought deep neural network (DNN) that is defined by restrictive Boltzmann machines (RBMs) \cite{Hinton2012} into HMM based models \cite{hinton2012deepSpeech,li2013hybrid,Miao2013ImprovingLC}. RBM based HMM is trained with a hierarchical scheme consisting of multiple-step unsupervised learning, formatting of a classification network and then supervised learning. The hierarchical procedure comes from the empirical expertise in this domain. To be more specific, the hierarchical learning scheme of RBM/DNN based HMM consists of: i) RBMs are trained one after the other in unsupervised fashion, and are stacked together as one deep neural network model, ii) then a final softmax layer is added to the stack of RBMs to represent the probability of a HMM state given a data input, iii) a discriminative training is performed for the final tuning of the model at the final stage.

@@

In this work, we propose a generative model based HMM, termed as GenHMM. Specifically, a generative model in our GenHMM is generator-mixed, where a generator is realized by a neural network to help the model gain high modeling capacity.
Our proposed model, GenHMM:
\begin{itemize}
\item has high modeling capacity of sequential data, due to the neural network based generators;
\item is easy to train. Training of GenHMM employs expectation maximization (EM) framework. Therefore, training a GenHMM is as easy as training a GMM-HMM model, while configuration of GenHMM is flexible;
  % \item has high modeling capability of sequence data, though its training and using are handy
\item is able to compute loglikelihood exactly and efficiently.
\end{itemize}
Instead of using softmax for probability representation, our GenHMM has tractability of exact loglikelihood of given sequential data, which is based on the change of variable formula. To make the loglikelihood computation efficient, neural network based generators of GenHMM are realized as flow models.

Our contributions in the paper are as follows:
\begin{itemize}
\item Proposing a neural network based HMM for sequential data modeling, i.e. GenHMM. GenHMM has the tractability of exact likelihood.
\item Designing practical algorithm for training GenHMM under EM framework. Stochastic gradient  search in batch fashion is embedded in this algorithm.
\item Giving convergence analysis for GenHMM under the proposed learning algorithm.
\item Verifying the proposed model on practical sequential data.
\end{itemize}

@@

Our framework is a HMM. A HMM $\bm{H}$ defined in a hypothesis space $\Hh$, i.e. $\bm{H} \in \Hh$, is capable to model time-span signal $\ubar{\bm{x}} = \left[ \bm{x}_1, \cdots, \bm{x}_T\right]^{\intercal}$, where $\bm{x}_t\in \RR^{N}$ is the $N$-dimensional signal at time $t$, $[\cdot]^{\intercal}$ denotes transpose, and $T$ denotes the time length\footnote{The length for  sequential data varies.}. We define the hypothesis set of HMM as $\Hh := \{\bm{H} | \bm{H}=\{\Ss, \bm{q}, \bm{A}, p(\bm{x}|{s}; \bm{\Phi}_{s})\}\}$, where
\begin{itemize}
\item $\Ss$ is the set of hidden states of $\bm{H}$.
\item $\bm{q} = \left[ q_1, q_2, \cdots, q_{|\Ss|}\right]^\intercal$ is the initial distribution of $\bm{H}$ with $|\Ss|$ as cardinality of $\Ss$. For $i \in \Ss$, $q_i = p(s_{1}=i;\bm{H})$. We use $s_t$ to denote the state $s$ at time $t$.
\item $\bm{A}$ matrix of size $|\Ss| \times |\Ss|$ is the transition matrix of $\bm{H}$. That is, $\forall i, j \in \Ss$,  $\bm{A}_{i,j} = p(s_{t+1}=j|s_{t}=i; \bm{H})$.
\item For a given hidden state $s$, the density function of the observable signal is $p({\bm{x}}|{s};\bm{\Phi}_{s})$, where $\bm{\Phi}_{s}$ is the parameter set that defines this probabilistic model. Denote $\bm{\Phi} = \left\{ \bm{\Phi}_{s}| s \in \Ss \right\}$.
\end{itemize}

@@

In this section, we introduce the neural network based state probabilistic model of our GenHMM.
Recall that $\bm{x}\in\mathbb{R}^N$.
The probabilistic model of GenHMM for each hidden state is a mixture of $K$ neural network based generators, where $K$ is a positive integer.
The probabilistic model of a state $s\in\Ss$ is then given by
\begin{equation}\label{eq:state-prob-model}
  p(\bm{x}| s; \bm{\Phi}_{s}) = \sum_{\kappa=1}^{K}\pi_{s, \kappa} p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa}),
\end{equation}
where $\kappa$ is a random variable following a categorical distribution, with probability $\pi_{s, \kappa} = p(\kappa | s; \bm{H})$.
Naturally $\sum_{\kappa = 1}^{K} \pi_{s, \kappa}= 1$. Denote $\bm{\pi}_{s} = [\pi_{s,1}, \pi_{s,2}, \cdots, \pi_{s,K}]^{\intercal}$. 
In \eqref{eq:state-prob-model}, $p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa})$ is defined as induced distribution by a generator $\bm{g}_{s,\kappa}: \RR^{N}\rightarrow\RR^{N}$, such that $\bm{x}=\bm{g}_{s, \kappa}(\bm{z})$, where $\bm{z}$ is a latent variable following a distribution with density function $p_{s,\kappa}(\bm{z})$. Generator $\bm{g}_{s,\kappa}$ is parameterized by $\bm{\theta}_{s, \kappa}$. Let us denote the collection of the parameter sets of generators for state $s$ as $\bm{\theta}_s = \left\{ \bm{\theta}_{s, \kappa}| \kappa = 1, 2, \cdots, K \right\}$. Assuming $\bm{g}_{s, \kappa}$ is invertible, by change of variable, we have
\begin{equation}\label{eq:changel-variable}
  p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa}) = p_{s,\kappa}(\bm{z})\bigg| \det\left( \pd{\bm{g}_{s,\kappa}(\bm{z})}{\bm{z}} \right)\bigg|^{-1}.
\end{equation}

@@

Since directly maximizing likelihood is not an option for our problem in \eqref{eq:ml-of-hmm}, we address this problem in expectation maximization (EM) framework. This divides our problem into  two iterative steps: i) using the joint posterior of hidden variable sequences $\ubm{s}$ and $\ubm{\kappa}$ to obtain an ``expected likelihood'' of the observable variable sequence $\ubm{x}$, i.e. the E-step; ii) maximizing the expected likelihood with regard to (w.r.t.) the model $\bm{H}$, i.e. the M-step. Assume model $\bm{H}$ is at a configuration of $\bm{H}^{\mathrm{old}}$, we formulate these two steps as follows:
\begin{itemize}
\item E-step: % the posterior probability of $\ubm{s}$:
  % \begin{equation}
  %   p(\ubm{s}|\ubm{x})
  % \end{equation}
  the expected likelihood function
  \begin{equation}\label{eq:em-q-funciton}
    \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) = \EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})}\left[ \log\,p(\ubm{x}, \ubm{s}, \ubm{\kappa}; \bm{H})\right],
  \end{equation}
  where $\EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})}\left[ \cdot\right]$ denotes the expectation operator by distribution $\hat{p}(\ubm{x})$ and $p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})$.
\item M-step: the maximization step
  \begin{equation}\label{eq:em-m-opt}
    \umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}}).
  \end{equation}
\end{itemize}

The problem \eqref{eq:em-m-opt} can be reformulated as
\begin{align}\label{eq:m-step-subs}
  &\umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) \nonumber \\
  =&\umax{\bm{q}}\Qq(\bm{q}; \bm{H}^{\mathrm{old}}) + \umax{\bm{A}}\Qq(\bm{A}; \bm{H}^{\mathrm{old}}) 
     + \umax{\bm{\Phi}}\Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}),
\end{align}
where
\begin{align}
  \Qq(\bm{q}; \bm{H}^{\mathrm{old}}) 
  % &=\EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})} \left[ \log\,p({s}_{1})  \right] \nonumber\\
    &= \EE_{\hat{p}(\ubm{x}),p(\ubm{s}| \ubm{x}; \bm{H}^{\mathrm{old}})} \left[ \log\,p({s}_{1};\bm{H})  \right], \label{eq:init-distribution-update}\\
  \Qq(\bm{A}; \bm{H}^{\mathrm{old}}) &=\EE_{\hat{p}(\ubm{x}),p(\ubm{s}| \ubm{x}; \bm{H}^{\mathrm{old}})}\hspace{-0.1cm}\left[ \sum_{t=1}^{T}\log\,p({s}_{t+1}|{s}_{t}; \bm{H}) \right], \label{eq:transition-update}\\
  \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}) &= \EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})} \left[ \log\,p(\ubm{x}, \ubm{\kappa}| \ubm{s}; \bm{H}) \right]. \label{eq:generative-model-update}
\end{align}

@@

To solve problem \eqref{opm:pi}, we formulate its Lagrange function as
\begin{equation}
  \Ll = \Qq(\bm{\Pi}; \bm{H}^{\mathrm{old}}) + \sum_{s=1}^{|\Ss|} \lambda_s\left( 1-  \sum_{\kappa=1}^{K}\pi_{s, \kappa}  \right).
\end{equation}
Solving $\pd{\Ll}{\pi_{s, \kappa}} = 0$ gives
\begin{equation}
  \pi_{s,\kappa} = \frac{1}{\lambda_s}\sum_{r=1}^{R} \sum_{t=1}^{{T}^{r}} p(\smtr{s}{t}{r}=s, \smtr{\kappa}{t}{r} =\kappa| \ubmr{x}{r}; \bm{H}^{\mathrm{old}}).
\end{equation}
With condition $\sum_{\kappa=1}^{K} \pi_{s, \kappa}= 1, \forall s = 1, 2, \cdots, |\Ss|$, we have
\begin{equation}
  \lambda_s = \sum_{\kappa=1}^{K}\sum_{r=1}^{R} \sum_{t=1}^{{T}^{(r)}} p(\smtr{s}{t}{r}=s, \smtr{\kappa}{t}{r} =\kappa | \ubmr{x}{r}; \bm{H}^{\mathrm{old}}).
\end{equation}
Then the solution to \eqref{opm:pi} is
\begin{equation}\label{eq:mix-latent-parameter-solution}
  \pi_{s, \kappa} = \frac{\sum_{r=1}^{R} \sum_{t=1}^{{T}^{r}} p(\smtr{s}{t}{r} =s, \smtr{\kappa}{t}{r}=\kappa | \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) }{\sum_{k =1}^{K}\sum_{r=1}^{R} \sum_{t=1}^{{T}^{r}-1} p(\smtr{s}{t}{r} =s, \smtr{\kappa}{t}{r}=k | \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) },
\end{equation}
where $p(s, \kappa | \ubm{x}; \bm{H}^{\mathrm{old}}) = p(s| \ubm{x}; \bm{H}^{\mathrm{old}}) p(\kappa | s, \ubm{x}; \bm{H}^{\mathrm{old}})$. $p(s| \ubm{x}; \bm{H}^{\mathrm{old}})$ can be computed by forward-backward algorithm, while $p(\kappa | s, \ubm{x}; \bm{H}^{\mathrm{old}})$ is given by \eqref{eq:kappa-posterior}.

@@

\subsubsection{Initial Probability Update}
The problem in \eqref{eq:init-distribution-update} can be reformulated as
\begin{align}
  \Qq(\bm{q}; \bm{H}^{\mathrm{old}})
  &=\frac{1}{R} \sum_{r=1}^{R}\sum_{\ubmr{s}{r}} {p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{1}{r};\bm{H}) \nonumber \\
  % & = \frac{1}{R}\sum_{r=1}^{R}\sum_{\smtr{s}{1}{r}=1}^{|\Ss|}\sum_{\smtr{s}{2}{r}=1}^{|\Ss|}\cdots \sum_{\smtr{s}{T^{r}}{r}}^{{|\Ss|}} {p(\smtr{s}{1}{r}, \smtr{s}{2}{r}, \cdots, \smtr{s}{T^{r}}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{1}{r}) \\
  & = \frac{1}{R}\sum_{r=1}^{R}\sum_{\smtr{s}{1}{r}=1}^{|\Ss|}{p(\smtr{s}{1}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{1}{r};\bm{H}).
\end{align}

@@

    \geq & \EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})}\left[ \log\,\frac{p(\ubm{x}, \ubm{s}, \ubm{\kappa}; \bm{H}^{\mathrm{old}})}{p(\ubm{s}, \ubm{\kappa}|\ubm{x}; \bm{H}^{\mathrm{old}})}\right] \nonumber \\

@@

\subsection{Algorithm of GenHMM}
\begin{algorithm}[H]
  \caption{Learning of GenHMM}\label{algo:genhmm}
  \begin{algorithmic}[1]
    \STATE {\bfseries Input:}{
      Empirical distribution $\hat{p}(\bm{x})$ of dataset\\
      Initializing $\bm{H}^{\mathrm{old}}, \bm{H} \in \Hh$ gives: \\
      $\bm{H}^{\mathrm{old}} = \{\Ss, \bm{q}^{\mathrm{old}}, A^{\mathrm{old}}, p(\bm{x}|s; \bm{\Phi}_{s}^{\mathrm{old}})\}$ \\
      $\bm{H} = \{\Ss, \bm{q}, A, p(\bm{x}|s; \bm{\Phi}_{s})\}$} \\
    \STATE $\bm{H}^{\mathrm{old}} \gets \bm{H}$
    \STATE Set learning rate $\eta$, neural network optimization batches $N$ per EM step
    \FOR { $\bm{H}$ not converge}
    \FOR {epoch $n < N$}
    \STATE Sample a batch of data $\left\{ \ubmr{x}{r} \right\}_{r=1}^{R_b}$ from dataset $\hat{p}(\ubm{x})$ with batch size $R_b$

@@

To summarize the learning solution in subsection~\ref{subsec:optmGenHMM}, we wrap our algorithm into pseudocode as shown in Algorithm~\ref{algo:genhmm}. We use Adam \cite{DBLP:journals/corr/KingmaB14} optimizer for optimization w.r.t. the parameters of generators in GenHMM. As shown from line $6$ to $12$ in Algorithm~\ref{algo:genhmm}, the batch-size stochastic gradient decent can be naturally embedded into the learning algorithm of GenHMM.

@@

Our experiments are performed for: i) standard classification tasks (Table~\ref{tab:acc-classification39f_a}, \ref{tab:acc-classification39f_b}, \ref{tab:acc-classification13f_a}, \ref{tab:acc-classification13f_b}), ii) classification under noise perturbation (table~\ref{tab:acc-classification39f_noise_snr}, \ref{tab:acc-classification39f_noise_type}). The criterion used to report the results includes accuracy, precision and F1 score.
In all experiments, generators $\left\{\bm{g}_{s,\kappa}|s\in \Ss, \kappa=1, 2, \cdots, K \right\}$ of GenHMM are implemented as flow models. Specifically, our generator structure follows that of a RealNVP described in \cite{2016arXiv160508803D}.
As discussed, the coupling layer shown in \eqref{eq-gl} maps a part of its input signal identically.
The implementation is such that layer $l+1$ would alternate the input signal order of layer $l$ such that no signal remains the same after two consecutive coupling layers.
We term such a pair of consecutive coupling layers as a \textit{flow block}.
In our experiments, each generator $\bm{g}_{s,\kappa}$ consists of four \textit{flow blocks}.
The density of samples in the latent space is defined as Normal, i.e. $p_{s,\kappa}(\bm{z})$ is the density function of standard Gaussian.

>>>>

Given the success of GMM-HMM, it is not efficient for modeling data in nonlinear manifold. Research attempts at training HMM with neural networks have been made to boost the modeling capacity of HMM. A successful work of this track has brought deep neural network (DNN) that is defined by restrictive Boltzmann machines (RBMs) \cite{Hinton2012} into HMM based models \cite{hinton2012deepSpeech,li2013hybrid,Miao2013ImprovingLC}. RBM based HMM is trained with a hierarchical scheme consisting of multiple steps of unsupervised learning, formatting of a classification network and then supervised learning. The hierarchical procedure comes from the empirical expertise in this domain. To be more specific, the hierarchical learning scheme of RBM/DNN based HMM consists of: i) RBMs are trained one after the other in unsupervised fashion, and are stacked together as one deep neural network model, ii) then a final softmax layer is added to the stack of RBMs to represent the probability of a HMM state given a data input, iii) a discriminative training is performed for the final tuning of the model at the final stage.

@@

In this work, we propose a generative model based HMM, termed as GenHMM. Specifically, a generative model in our GenHMM is generator-mixed, where a generator is realized by a neural network to help the model gain high modeling capacity.
Our proposed model, GenHMM,
\begin{itemize}
\item has high modeling capacity of sequential data, due to the neural network based generators;
\item is easy to train. Training of GenHMM employs expectation maximization (EM) framework. Therefore, training a GenHMM is as easy as training a GMM-HMM model, while configuration of GenHMM is flexible;
  % \item has high modeling capability of sequence data, though its training and using are handy
\item is able to compute loglikelihood exactly and efficiently.
\end{itemize}
Instead of using softmax for probability representation, our GenHMM has tractability of exact loglikelihood of given sequential data, which is based on the change of variable formula. To make the loglikelihood computation efficient, neural network based generators of GenHMM are realized as flow models.

Our contributions in the paper are as follows.
\begin{itemize}
\item Proposing a neural network based HMM for sequential data modeling, i.e. GenHMM. GenHMM has the tractability of exact likelihood.
\item Designing practical algorithm for training GenHMM under EM framework. Stochastic gradient  search in batch fashion is embedded in this algorithm.
\item Giving convergence analysis for GenHMM under the proposed learning algorithm.
\item Verifying the proposed model on practical sequential data.
\end{itemize}

@@

Our framework is a HMM. A HMM $\bm{H}$ defined in a hypothesis space $\Hh$, i.e. $\bm{H} \in \Hh$, is capable to model time-span signal $\ubar{\bm{x}} = \left[ \bm{x}_1, \cdots, \bm{x}_T\right]^{\intercal}$, where $\bm{x}_t\in \RR^{N}$ is the $N$-dimensional signal at time $t$, $[\cdot]^{\intercal}$ denotes transpose, and $T$ denotes the time length\footnote{The length for  sequential data varies.}. We define the hypothesis set of HMM as $\Hh := \{\bm{H} | \bm{H}=\{\Ss, \bm{q}, \bm{A}, p(\bm{x}|{s}; \bm{\Phi}_{s})\}\}$, where
\begin{itemize}
\item $\Ss$ is the set of hidden states of $\bm{H}$.
\item $\bm{q} = \left[ q_1, q_2, \cdots, q_{|\Ss|}\right]^\intercal$ is the initial state distribution of $\bm{H}$ with $|\Ss|$ as cardinality of $\Ss$. For $i \in \Ss$, $q_i = p(s_{1}=i;\bm{H})$. We use $s_t$ to denote the state $s$ at time $t$.
\item $\bm{A}$ matrix of size $|\Ss| \times |\Ss|$ is the transition matrix of states in $\bm{H}$. That is, $\forall i, j \in \Ss$,  $\bm{A}_{i,j} = p(s_{t+1}=j|s_{t}=i; \bm{H})$.
\item For a given hidden state $s$, the density function of the observable signal is $p({\bm{x}}|{s};\bm{\Phi}_{s})$, where $\bm{\Phi}_{s}$ is the parameter set that defines this probabilistic model. Denote $\bm{\Phi} = \left\{ \bm{\Phi}_{s}| s \in \Ss \right\}$.
\end{itemize}

@@

In this section, we introduce the neural network based state probabilistic model of our GenHMM.
Recall that $\bm{x}\in\mathbb{R}^N$. Subscript is omitted when it does not cause ambiguity.
The probabilistic model of GenHMM for each hidden state is a mixture of $K$ neural network based generators, where $K$ is a positive integer.
The probabilistic model of a state $s\in\Ss$ is then given by
\begin{equation}\label{eq:state-prob-model}
  p(\bm{x}| s; \bm{\Phi}_{s}) = \sum_{\kappa=1}^{K}\pi_{s, \kappa} p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa}),
\end{equation}
where $\kappa$ is a random variable following a categorical distribution, with probability $\pi_{s, \kappa} = p(\kappa | s; \bm{H})$.
Naturally $\sum_{\kappa = 1}^{K} \pi_{s, \kappa}= 1$. Denote $\bm{\pi}_{s} = [\pi_{s,1}, \pi_{s,2}, \cdots, \pi_{s,K}]^{\intercal}$. 
In \eqref{eq:state-prob-model}, $p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa})$ is defined as induced distribution by a generator $\bm{g}_{s,\kappa}: \RR^{N}\rightarrow\RR^{N}$, such that $\bm{x}=\bm{g}_{s, \kappa}(\bm{z})$, where $\bm{z}$ is a latent variable following a distribution with density function $p_{s,\kappa}(\bm{z})$. Generator $\bm{g}_{s,\kappa}$ is parameterized by $\bm{\theta}_{s, \kappa}$. Let us denote the collection of the parameter sets of generators for state $s$ as $\bm{\theta}_s = \left\{ \bm{\theta}_{s, \kappa}| \kappa = 1, 2, \cdots, K \right\}$. Assuming $\bm{g}_{s, \kappa}$ is invertible, by change of variable, we have
\begin{equation}\label{eq:changel-variable}
  p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa}) = p_{s,\kappa}(\bm{z})\bigg| \det\left( \pd{\bm{g}_{s,\kappa}(\bm{z})}{\bm{z}} \right)\bigg|^{-1}.
\end{equation}

@@

Since directly maximizing likelihood is not an option for our problem in \eqref{eq:ml-of-hmm}, we address this problem in expectation maximization (EM) framework. This divides our problem into  two iterative steps: i) using the joint posterior of hidden variable sequences $\ubm{s}$ and $\ubm{\kappa}$ to obtain an ``expected likelihood'' of the observable variable sequence $\ubm{x}$, i.e. the E-step; ii) maximizing the expected likelihood with regard to (w.r.t.) the model $\bm{H}$, i.e. the M-step. Assume model $\bm{H}$ is at a configuration of $\bm{H}^{\mathrm{old}}$, we formulate these two steps as follows.
\begin{itemize}
\item E-step: % the posterior probability of $\ubm{s}$:
  % \begin{equation}
  %   p(\ubm{s}|\ubm{x})
  % \end{equation}
  the expected likelihood function
  \begin{equation}\label{eq:em-q-funciton}
    \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) = \EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})}\left[ \log\,p(\ubm{x}, \ubm{s}, \ubm{\kappa}; \bm{H})\right],
  \end{equation}
  where $\EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})}\left[ \cdot\right]$ denotes the expectation operator by distribution $\hat{p}(\ubm{x})$ and $p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})$.
\item M-step: the maximization step
  \begin{equation}\label{eq:em-m-opt}
    \umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}}).
  \end{equation}
\end{itemize}

The problem \eqref{eq:em-m-opt} can be reformulated as
\begin{align}\label{eq:m-step-subs}
  &\umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) \nonumber \\
  =&\umax{\bm{q}}\Qq(\bm{q}; \bm{H}^{\mathrm{old}}) + \umax{\bm{A}}\Qq(\bm{A}; \bm{H}^{\mathrm{old}}) 
     + \umax{\bm{\Phi}}\Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}),
\end{align}
where the decomposed optimization problems are
\begin{align}
  \Qq(\bm{q}; \bm{H}^{\mathrm{old}}) 
  % &=\EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})} \left[ \log\,p({s}_{1})  \right] \nonumber\\
    &= \EE_{\hat{p}(\ubm{x}),p(\ubm{s}| \ubm{x}; \bm{H}^{\mathrm{old}})} \left[ \log\,p({s}_{1};\bm{H})  \right], \label{eq:init-distribution-update}\\
  \Qq(\bm{A}; \bm{H}^{\mathrm{old}}) &=\EE_{\hat{p}(\ubm{x}),p(\ubm{s}| \ubm{x}; \bm{H}^{\mathrm{old}})}\hspace{-0.1cm}\left[ \sum_{t=1}^{T-1}\log\,p({s}_{t+1}|{s}_{t}; \bm{H}) \right], \label{eq:transition-update}\\
  \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}) &= \EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})} \left[ \log\,p(\ubm{x}, \ubm{\kappa}| \ubm{s}; \bm{H}) \right]. \label{eq:generative-model-update}
\end{align}

@@

To solve problem \eqref{opm:pi}, we formulate its Lagrange function as
\begin{equation}
  \Ll = \Qq(\bm{\Pi}; \bm{H}^{\mathrm{old}}) + \sum_{s=1}^{|\Ss|} \lambda_s\left( 1-  \sum_{\kappa=1}^{K}\pi_{s, \kappa}  \right).
\end{equation}
Solving $\pd{\Ll}{\pi_{s, \kappa}} = 0$ gives
\begin{equation}
  \pi_{s,\kappa} = \frac{1}{\lambda_s}\sum_{r=1}^{R} \sum_{t=1}^{{T}^{r}} p(\smtr{s}{t}{r}=s, \smtr{\kappa}{t}{r} =\kappa| \ubmr{x}{r}; \bm{H}^{\mathrm{old}}).
\end{equation}
With condition $\sum_{\kappa=1}^{K} \pi_{s, \kappa}= 1, \forall s = 1, 2, \cdots, |\Ss|$, we have
\begin{equation}
  \lambda_s = \sum_{\kappa=1}^{K}\sum_{r=1}^{R} \sum_{t=1}^{{T}^{r}} p(\smtr{s}{t}{r}=s, \smtr{\kappa}{t}{r} =\kappa | \ubmr{x}{r}; \bm{H}^{\mathrm{old}}).
\end{equation}
Then the solution to \eqref{opm:pi} is
\begin{equation}\label{eq:mix-latent-parameter-solution}
  \pi_{s, \kappa} = \frac{\sum_{r=1}^{R} \sum_{t=1}^{{T}^{r}} p(\smtr{s}{t}{r} =s, \smtr{\kappa}{t}{r}=\kappa | \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) }{\sum_{k =1}^{K}\sum_{r=1}^{R} \sum_{t=1}^{{T}^{r}} p(\smtr{s}{t}{r} =s, \smtr{\kappa}{t}{r}=k | \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) },
\end{equation}
where
\begin{equation}
  p(s, \kappa | \ubm{x}; \bm{H}^{\mathrm{old}}) = p(s| \ubm{x}; \bm{H}^{\mathrm{old}}) p(\kappa | s, \ubm{x}; \bm{H}^{\mathrm{old}}).
\end{equation}
Here $p(s| \ubm{x}; \bm{H}^{\mathrm{old}})$ can be computed by forward-backward algorithm, while $p(\kappa | s, \ubm{x}; \bm{H}^{\mathrm{old}})$ is given by \eqref{eq:kappa-posterior}.

@@

\subsubsection{Initial Probability Update}
The problem in \eqref{eq:init-distribution-update} can be reformulated as
\begin{align}
  &\Qq(\bm{q}; \bm{H}^{\mathrm{old}}) \nonumber \\
  &=\frac{1}{R} \sum_{r=1}^{R}\sum_{\ubmr{s}{r}} {p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{1}{r};\bm{H}) \nonumber \\
  & = \frac{1}{R}\sum_{r=1}^{R}\sum_{\smtr{s}{1}{r}=1}^{|\Ss|}\sum_{\smtr{s}{2}{r}=1}^{|\Ss|}\cdots \sum_{\smtr{s}{T^{r}}{r}}^{{|\Ss|}} {p(\smtr{s}{1}{r}, \smtr{s}{2}{r}, \cdots, \smtr{s}{T^{r}}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{1}{r}) \nonumber\\
  & = \frac{1}{R}\sum_{r=1}^{R}\sum_{\smtr{s}{1}{r}=1}^{|\Ss|}{p(\smtr{s}{1}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{1}{r};\bm{H}).
\end{align}

@@

    \geq & \EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x};
           \bm{H}^{\mathrm{old}})}\left[ \log\,\frac{p(\ubm{x},
           \ubm{s}, \ubm{\kappa}; \bm{H}^{\mathrm{old}})}{p(\ubm{s},
           \ubm{\kappa}|\ubm{x}; \bm{H}^{\mathrm{old}})}\right]
           \nonumber \\
    &+ \EE_{\hat{p}(\ubm{x})}\left[ KL(p(\ubm{s}, \ubm{\kappa}|\ubm{x}; \bm{H}^{\mathrm{old}})\|p(\ubm{s}, \ubm{\kappa}|\ubm{x}; \bm{H}^{\mathrm{old}})) \right]
      \nonumber \\

@@

\subsection{Algorithm of GenHMM}
\begin{algorithm}[H]
  \caption{Learning of GenHMM}\label{algo:genhmm}
  \begin{algorithmic}[1]
    \STATE {\bfseries Input:}{
      Empirical distribution $\hat{p}(\bm{x})$ of dataset}\\
    \STATE Initializing $\bm{H}^{\mathrm{old}}, \bm{H} \in \Hh$ gives: \\
    $\bm{H}^{\mathrm{old}} = \{\Ss, \bm{q}^{\mathrm{old}}, A^{\mathrm{old}}, p(\bm{x}|s; \bm{\Phi}_{s}^{\mathrm{old}})\}$, \\
    $\bm{H} = \{\Ss, \bm{q}, A, p(\bm{x}|s; \bm{\Phi}_{s})\}$, \\
    in which generators $\left\{\bm{g}_{s,\kappa}|s\in \Ss, \kappa=1,
      2, \cdots, K \right\}$ are all initialized randomly.
    \STATE $\bm{H}^{\mathrm{old}} \gets \bm{H}$
    \STATE Set learning rate $\eta$, neural network optimization batches $N$ per EM step
    \FOR { $\bm{H}$ not converge}
    \FOR {epoch $n < N$}
    \STATE Sample a batch of data $\left\{ \ubmr{x}{r} \right\}_{r=1}^{R_b}$ from dataset $\hat{p}(\ubm{x})$ with batch size $R_b$

@@

To summarize the learning solution in subsection~\ref{subsec:optmGenHMM}, we wrap our algorithm into pseudocode as shown in Algorithm~\ref{algo:genhmm}. We use Adam \cite{DBLP:journals/corr/KingmaB14} optimizer for optimization w.r.t. the parameters of generators in GenHMM. As shown from line $6$ to $10$ in Algorithm~\ref{algo:genhmm}, the batch-size stochastic gradient decent can be naturally embedded into the learning algorithm of GenHMM.

@@

Our experiments are performed for: i) standard classification tasks (Table~\ref{tab:acc-classification39f_a}, \ref{tab:acc-classification39f_b}, \ref{tab:acc-classification13f_a}, \ref{tab:acc-classification13f_b}), ii) classification under noise perturbation (table~\ref{tab:acc-classification39f_noise_snr}, \ref{tab:acc-classification39f_noise_type}). The criterion used to report the results includes accuracy, precision and F1 scores.
In all experiments, generators $\left\{\bm{g}_{s,\kappa}|s\in \Ss, \kappa=1, 2, \cdots, K \right\}$ of GenHMM are implemented as flow models. Specifically, our generator structure follows that of a RealNVP described in \cite{2016arXiv160508803D}.
As discussed, the coupling layer shown in \eqref{eq-gl} maps a part of its input signal identically.
The implementation is such that layer $l+1$ would alternate the input signal order of layer $l$ such that no signal remains the same after two consecutive coupling layers.
We term such a pair of consecutive coupling layers as a \textit{flow block}.
In our experiments, each generator $\bm{g}_{s,\kappa}$ consists of four \textit{flow blocks}.
The density of samples in the latent space is defined as Normal,
i.e. $p_{s,\kappa}(\bm{z})$ is the density function of standard
Gaussian. The configuration for each generator is shown as Table~\ref{table:generator-setting}.

\begin{table}
  \caption{Configuration of generators in Experiments}\label{table:generator-setting}
  \centering
  \begin{tabular}{l|c}
    \toprule
    \makecell{Latent distribution $p_{s,\kappa}(\bm{z})$
    \\$s\in \Ss, \kappa=1, 2, \cdots, K $} & Standard Gaussian \\
    \hline
    Number of flow blocks & $4$ \\
    \hline
    Non-linear mapping $\bm{m}_a$, $\bm{m}_b$ & \makecell{Multiple
                                                layer perception, \\
    $3$ layers and with hidden dimension $24$}\\
    \bottomrule
  \end{tabular}
\end{table}
